Open archaeology, open source?

### 2024-04-09 analysis/_02-background.qmd
Additional changes based on reviewer comments.

The idea that scientists should generally contribute to a public domain of knowledge without profit motive has led to open science being heralded as revolutionary, community-oriented, and anti-capitalist means of production. However, while open science does have the _potential_ to effect radical change, this is not a given. The social and institutional contexts in which we do science is firmly embedded within capitalist and neoliberal power structures that reward individualistic competition and do little to actually encourage equitable and accessible research practices, and as such, make it difficult to fully embrace open science ideals [@mirowski2018]. Moreover, the open science movement, which is dominated by STEM disciplines, prioritizes a grossly simplified and asocial notion of what science is and entails. Namely, it considers science as the accumulation and assembly of a species-level understanding of the world, which is not held by any one individual but is stored in seemingly value-neutral and disembodied media, facts and observations. This is manifested by information and communication technologies that host files, document processes, facilitate co-working opportunities, and perform automated processes (e.g. digital repositories, reproducible notebooks, automated evaluation of research findings, systems for ensuring stable references to consistent identifiers, etc). However, these systems have become so emblematic of open science that the use of these tools and resources designed to _support_ open science is often mistaken for _actually doing_ open science.

>>>>

The idea that scientists should generally contribute to a public domain of knowledge without profit motive has led to open science being heralded as revolutionary, community-oriented, and anti-capitalist means of production. However, while open science does have the _potential_ to effect radical change, this is not a given. The social and institutional contexts in which we do science is firmly embedded within capitalist and neoliberal power structures that reward individualistic competition and do little to actually encourage equitable and accessible research practices, and as such, make it difficult to fully embrace open science ideals [@mirowski2018]. Moreover, the open science movement, which is dominated by STEM disciplines, prioritizes a grossly simplified and asocial notion of what science is and entails. Namely, it considers science as the accumulation and assembly of a species-level understanding of the world, which is not held by any one individual but is stored in seemingly value-neutral and disembodied media, facts and observations. This is manifested by information and communication technologies that host files, document processes, facilitate co-working opportunities, and perform automated processes (e.g. digital repositories, reproducible notebooks, automated evaluation of research findings, systems for ensuring stable references to consistent identifiers, etc). However, these systems have become so emblematic of open science that the use of these tools and resources designed to _support_ open science is often mistaken for _actually doing_ open science [@leonelli2023: 23-24].

### 2024-04-09 analysis/_07-conclusion.qmd
Additional changes based on reviewer comments.

These findings call into question the notion that archaeologists benefit from the positive outcomes that are commonly argued to be the natural results of open source development models -- namely, greater degrees of extensibility and participatory action.
While opening the source code may facilitate these positive outcomes as necessary preconditional factors, we argue that this only amounts to establishing the _potential_ for people to put these values into practice. 
Moreover, we argue that the objectives and circumstances that frame archaeological practice significantly influence how far archaeologists (and academics in general) are willing to push for these values, and limit the ability for archaeologists to do open source in ways that resemble more mainstream open source projects.
For instance, successful open source projects like the Linux kernel, openSSL, or the Firefox web browser are driven by collective and popular interest in ensuring that code remains functional, and the code base is therefore constantly in flux and bears an accumulating list of contributing members.
This differs from the organizational principles that govern much archaeological work, namely where a director or directors (of a field project, research group, etc.) sets the goals and orientation of the group and commissions and manages other actors accordingly. 
Moreover, archaeological projects ultimately seek to produce stable textual outcomes bearing clear delineation of authorship and that require no upkeep whatsoever. 
Sustaining an open source project is simply not compatible with the factors that currently drive the momentum behind archaeological work.

>>>>

These findings call into question the notion that archaeologists benefit from the positive outcomes that are commonly argued to be the natural results of open source development models -- namely, greater degrees of extensibility and participatory action.
While opening the source code may facilitate these positive outcomes as necessary preconditional factors, we argue that this only amounts to establishing the _potential_ for people to put these values into practice.
We argue that the objectives and circumstances that frame archaeological practice significantly influence how far archaeologists (and academics in general) are willing to push for these values, and limit the ability for archaeologists to do open source in ways that resemble more mainstream open source projects.
For instance, successful open source projects like the Linux kernel, openSSL, or the Firefox web browser are driven by collective and popular interest in ensuring that code remains functional, and the code base is therefore constantly in flux and bears an accumulating list of contributing members.
This differs from the organizational principles that govern much archaeological work, namely where a director or directors (of a field project, research group, etc.) sets the goals and orientation of the group and commissions and manages other actors accordingly. 
Moreover, archaeological projects ultimately seek to produce stable textual outcomes bearing clear delineation of authorship and that require no upkeep whatsoever. 
Sustaining an open source project is simply not compatible with the factors that currently drive the momentum behind archaeological work.
